---
title: "Practical Work"
subtitle: "Lifetime Data Analysis"
author: "Rodrigo Arriaza, Alexander J Ohrt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_book:
    keep_tex: true
    toc: false
    number_sections: false
params:
  show_code = TRUE
---

```{r setup, include=FALSE}
list(rm = ls())
knitr::opts_chunk$set(echo = params$show_code, comment = "#>", eval = TRUE, warning = F)
library(tidyverse)
library(corrplot)
library(GGally)
library(survival)
library(FHtest) 
library(GofCens)
```

# Introduction

We are given a data set on sexually transmitted diseases (STDs). This is data from a study about gonorrhea and chlamydia in 877 women. The objective with this practical work is to study possible risk factors for a reinfection with gonorrhea or chlamydia in women who have suffered one of both infections previously. The variables of interest are sociodemographic variables or those related to sexual practice. We have a lot of variables at our disposal, as can be seen from the data set below. We have chosen to use the following:

  * Age: The age of the woman. 
  * NumPartners: The number of partners during the last 30 days. 
  * CondomUse: Use of condoms (1: always, 2: once in a while, 3: never)
  * YearsSchool: Years of schooling. 
  * InitInfect: Initial infection (1: Gonorrhea, 2: Chlamydia, 3: both)
  * InvVagAtExam: Involvement vagina at exam (1: yes; 0: no).
  * DischargeExam: Discharge at exam (1: yes; 0: no)

The first two were chosen based on results from a [study](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1744639/) on gonorrhea reinfection in heterosexual STD clinic attendees. The study concluded that increased reinfection risk (of gonorrhea) was associated with younger age and a greater number of recent sex partners, among other risk factors. Moreover, the authors concluded that any type of condom use was a risk factor for reinfection with gonorrhea in women. However by using statistical analysis we found that the Age of the woman and the ethnicity (a variable that was not included in the aforementioned study) are not statistically significant (Necessary to mention Ethnicity here, if it is not a part of the study?). Q: Should perhaps describe what we have done and why it has been done? I am going to ask Klaus about what he thinks about it also I think, just to see what he says. 

Another [publication](https://policylab.chop.edu/sites/default/files/pdf/publications/Preventing_Chlamydia_Gonorrhea_Reinfection_through_Increased_Use_of_EPT.pdf) reports that, on average, 14\% of women with clamydia and 12\% of women with gonorrhea get reinfected, with younger women at higher risk. Moreover, they state that many adolescents treated for infection of one of the two STDs are reinfected within three to six months, usually because of resumed sexual contact with an untreated partner. Thus, the marital status might be interesting to analyse. However, this is not added, because, as seen in the exploratory data analysis below, the ages are low, which should mean that the amount in each level of `MaritalStatus` is very skewed towards single. This can be seen in the table below as well. NOT SURE IF WE SHOULD KEEP THIS, DEPENDS ON THE AMOUNT OF ROOM I GUESS. can leave it until the end. 

[This meta-analysis](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2094865/) reports that the relationship between race, socioeconomic status (SES) and chlamydial infection is not clear. It concludes that SES was not associated with chlamydia infection, where they tested for several variables, where level of parent's education was one of them. Either way, we think it might be interesting to see if the years of schooling of the women have any impact on chlamydia reinfection and as is shown below it showed to be statistically significant during the exploratory analysis. Q: I also think that this would be a reason to look at race also (combined with the first study, where they only had afro-americans. Perhaps we can try to mix the variables chosen based on studies and on the statistical method? (that is; if this statistical method is something we can trust))


```{r}
std_data <- read.table("STD_onlydata.txt")
colnames(std_data) <- variable.names <- c("ObsNum", "Ethnicity", "MaritalStatus", "Age", "YearsSchool", "InitInfect", "NumPartners", "OralSex12m", "OralSex30d", "RectalSex12m", "RectalSex30d", "AbPain", "SignDischarge","SignDysuria","CondomUse","SignItch","SignLesion","SignRash","SignLymph","InvVagAtExam","DischargeExam","AbnormNodeExam", "Reinfection", "TimeUntilReinf")

std_data[, variable.names[-c(1, 4, 5, 7,23,24)]] <- lapply(std_data[, variable.names[-c(1,4,5,7,23,24)]],factor)
std_data$cens <- rep(1, length = dim(std_data)[[1]]) # All uncensored, thus cens = 1 for all rows. 

variables.chosen <- c("Age", "NumPartners", "CondomUse", "Ethnicity") # ETC!
continuous.variables <- unlist(lapply(std_data, is.numeric))
continuous.variables <- continuous.variables[-length(continuous.variables)] # Removing the cens-value. 
```

Naturally, the categorical variable which states if the woman is reinfected or not (`Reinfection`) will be used as a dependent variable in the analysis and the time until reinfection since the more time a subject is under study, the greater the risk of the event reocurring.

# Descriptive Analysis

## Variable selection 

Q: What is the offset(log(TimeUntilReInf))?

```{r variable-select}
std_data$Ethnicity <- factor(std_data$Ethnicity)
nb.model <- MASS::glm.nb(Reinfection ~ Ethnicity + MaritalStatus + Age + YearsSchool + InitInfect +
                           NumPartners +OralSex12m + OralSex30d + RectalSex12m + RectalSex30d + AbPain +
               SignDischarge + SignDysuria+ CondomUse + SignItch + SignLesion + SignRash +
               SignLymph + InvVagAtExam + DischargeExam + AbnormNodeExam + offset(log(TimeUntilReinf)),
             data=std_data)
s <- summary(nb.model)
k <- knitr::kable(s$coefficients, caption = 'Variables statistical significance')
kableExtra::row_spec(k, c(6:9,14,18,23,24), color='white', background = 'cyan')
```


```{r, include=F}
# Find a way to print it nicely!
knitr::kable(summary(std_data))
```


```{r, include=F}
hist(std_data[, "Age"], breaks = 100)
```

```{r, include=F}
knitr::kable(table(std_data[, "MaritalStatus"]), col.names = c("Marital Status", "Frequency"), caption = "Marital Status Frequency in each Level.")
knitr::kable(table(std_data[, "SignDischarge"]), col.names = c("Sign of Discharge", "Frequency"), caption = "Sign of Discharge Frequency in each Level.")
knitr::kable(table(std_data[, "DischargeExam"]), col.names = c("Discharge at Exam", "Frequency"), caption = "Discharge at Exam Frequency in each Level.")
# Some more tables?
```

[//]: # Pairs plot of the chosen variables. **Nothing really interested to see here perhaps?**

```{r, include=F}
ggpairs(std_data[,continuous.variables])
corrplot(cor(std_data[, continuous.variables]))
```
Note that age and years of schooling are somewhat correlated. **Could do some more EDA probably, and should remove some of this also.** 

# Nonparametric Analysis

The survival curve is estimated by means of Kaplan-Meier and plotted below. The curve below shows the general survival in the data set, i.e. not depending on any specific explanatory variable. 

```{r}
s1 <- with(std_data, Surv(TimeUntilReinf, cens) ~ 1)
s1fit <- survfit(s1)
#plot(s1fit)

par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(s1fit, col = 2, xlab = "Time to reinfection [days]",
     ylab = expression(bolditalic(hat(S)(t))),
     lty = 1, lwd = 3, yaxs = "i", xaxs = "i", bty = "n",
     conf.int = F)
title("Survival Function")
```

```{r}
s1.CondomUse <- with(std_data, Surv(TimeUntilReinf, cens) ~ CondomUse)
s1fit.CondomUse <- survfit(s1.CondomUse)
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(s1fit.CondomUse, col = c(1,2,3), xlab = "Time to reinfection [days]",
     ylab = expression(bolditalic(hat(S)(t))),
     yaxs = "i", xaxs = "i", bty = "n", lty = c(1,2,3), lwd = rep(3, length.out = 3),
     conf.int = F)
legend("topright", legend = levels(std_data$CondomUse), title = "Condom Use",
       bty = "n", col = c(1,2,3), lty = c(1,2,3), lwd = rep(3, length.out = 3))
title("Survival Function as Function of Condom Use")
```

```{r}
s1.Ethnicity <- with(std_data, Surv(TimeUntilReinf, cens) ~ Ethnicity)
s1fit.Ethnicity <- survfit(s1.Ethnicity)
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(s1fit.Ethnicity, col = c(1,2), xlab = "Time to reinfection [days]",
     ylab = expression(bolditalic(hat(S)(t))),
     yaxs = "i", xaxs = "i", bty = "n", lty = c(1,2), lwd = rep(3, length.out = 2),
     conf.int = F)
legend("topright", legend = levels(std_data$Ethnicity), title = "Ethnicity",
       bty = "n", col = c(1,2), lty = c(1,2), lwd = rep(3, length.out = 2))
title("Survival Function as Function of Ethnicity")
```

```{r}
s1.SignDischarge <- with(std_data, Surv(TimeUntilReinf, cens) ~ SignDischarge)
s1fit.SignDischarge <- survfit(s1.SignDischarge)
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(s1fit.SignDischarge, col = c(1,2), xlab = "Time to reinfection [days]",
     ylab = expression(bolditalic(hat(S)(t))),
     yaxs = "i", xaxs = "i", bty = "n", lty = c(1,2), lwd = rep(3, length.out = 2),
     conf.int = F)
legend("topright", legend = levels(std_data$SignDischarge), title = "Sign of Discharge",
       bty = "n", col = c(1,2), lty = c(1,2), lwd = rep(3, length.out = 2))
title("Survival Function as Function of Sign of Discharge")
```


```{r}
s1.NumPartners <- with(std_data, Surv(TimeUntilReinf, cens) ~ NumPartners)
s1fit.NumPartners <- survfit(s1.NumPartners)
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(s1fit.NumPartners, col = c(1,2,3,4,5,6,7,8,9), xlab = "Time to reinfection [days]",
     ylab = expression(bolditalic(hat(S)(t))),
     yaxs = "i", xaxs = "i", bty = "n", lty = c(1,2,3,4,5,6,7,8,9), lwd = rep(3, length.out = 9),
     conf.int = F)
legend("topright", legend = levels(as.factor(std_data$NumPartners)), title = "Number of Partners",
       bty = "n", col = c(1,2,3,4,5,6,7,8,9), lty = c(1,2,3,4,5,6,7,8,9), lwd = rep(3, length.out = 9))
title("Survival Function as Function of Number of Partners")
```


The median survival time is 247 days, as can be seen below. 

```{r}
s1fit
```


Comparison of survival functions by means of nonparametric tests, such as the logrank test. The curve on ethnicity could be interesting to test (the survivals cross though!)

Below, logrank test for all curves plotted above are done. 

```{r}
FHtestrcc(s1.CondomUse)
FHtestrcc(s1.Ethnicity)
FHtestrcc(s1.SignDischarge)
FHtestrcc(s1.NumPartners)
```

# Fit of a parametric survival model

Fit a Weibull, log-logistic or lognormal model. Justify the theoretical choice we are taking!!

```{r}
# I am testing all three below, since I am not sure which will be better.

# Below we have the null models. 
s2 <- with(std_data, Surv(TimeUntilReinf, cens))
loglo <- survreg(s2 ~ 1, dist = "loglo")  
summary(loglo)
weibull <- survreg(s2 ~ 1)
summary(weibull)
lognormal <- survreg(s2 ~ 1, dist = "lognormal")
summary(lognormal)

# Below we add some covariates to the models. INCLUDE ALL OF THEM PERHAPS! (OR TEST WITH DIFFERENT ONES!)
# First we add Ethnicity  e.g.
loglo.Ethnicity <- survreg(s2 ~ Ethnicity, dist = "loglo", data = std_data)  
summary(loglo.Ethnicity)
weibull.Ethnicity <- survreg(s2 ~ Ethnicity, data = std_data)
summary(weibull.Ethnicity)
lognormal.Ethnicity <- survreg(s2 ~ Ethnicity, dist = "lognormal", data = std_data)
summary(lognormal.Ethnicity)


loglo.full <- survreg(s2 ~ Ethnicity + Age + NumPartners + CondomUse + YearsSchool + SignDischarge, data = std_data, dist = "loglo")
summary(loglo.full)

loglo.pred <- predict(loglo.full, type = "linear")
resids.loglo <- (log(std_data$TimeUntilReinf) - loglo.pred) / loglo.full$scale

weibull.full <- update(loglo.full, dist = "weibull")
summary(weibull.full)

weibull.pred <- predict(weibull.full, type = "linear")
resids.weibull <- (log(std_data$TimeUntilReinf) - weibull.pred) / weibull.full$scale

lognormal.full <- update(loglo.full, dist = "lognormal")
summary(lognormal.full)

lognormal.pred <- predict(lognormal.full, type = "linear")
resids.logno <- (log(std_data$TimeUntilReinf) - lognormal.pred) / lognormal.full$scale

# Model checking (using residuals etc, as done in lab 6).
# Log-logistic
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(survfit(Surv(resids.loglo, std_data$cens) ~ 1), col = c(1,2,2), xlab = "Residuals",
     ylab = expression(bolditalic(hat(S)(t))),
     lty = 1, lwd = 3, yaxs = "i", xaxs = "i", bty = "n")
title("Residuals of the log-logistic Regression Model")
curve(plogis(x, lower.tail = F), from = min(resids.loglo), to = max(resids.loglo), col = 3, lwd = 3,
      add = TRUE)
legend("bottomleft", c("KM estimate", "95% - CI", "Stand. Logistic Distribution"),
       col = c(1, 2, 3), lty = c(1, 2, 1), lwd = 3, bty = "n")

# Weibull
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(survfit(Surv(resids.weibull, std_data$cens) ~ 1), col = c(1,2,2), xlab = "Residuals",
     ylab = expression(bolditalic(hat(S)(t))),
     lty = 1, lwd = 3, yaxs = "i", xaxs = "i", bty = "n")
title("Residuals of the Weibull Regression Model")
survgumb <- function(x) {
  return(exp(-exp(x)))
}
curve(survgumb, from = min(resids.weibull), to = max(resids.weibull), col = 3, lwd = 3,
      add = TRUE)
legend("bottomleft", c("KM estimate", "95% - CI", "Stand. Gumbel Distribution"),
       col = c(1, 2, 3), lty = c(1, 2, 1), lwd = 3, bty = "n")

# Log-normal
par(font = 2, font.axis = 2, font.lab = 2, las = 1, mar = c(5, 5, 4, 2))
plot(survfit(Surv(resids.logno, std_data$cens) ~ 1), col = c(1,2,2), xlab = "Residuals",
     ylab = expression(bolditalic(hat(S)(t))),
     lty = 1, lwd = 3, yaxs = "i", xaxs = "i", bty = "n")
title("Residuals of the Lognormal Regression Model")
curve(pnorm(x, lower.tail = F), from = min(resids.logno), to = max(resids.logno), col = 3, lwd = 3,
      add = TRUE)
legend("bottomleft", c("KM estimate", "95% - CI", "Stand. Normal Distribution"),
       col = c(1, 2, 3), lty = c(1, 2, 1), lwd = 3, bty = "n")
```

I would say that the Weibull looks the best based on the three residual plots that can be seen above. **But perhaps they should be tested in some other ways as well??**

Goodness of fit of the parametric models is done in lab6! These things can be used to choose which of the three distributions should be used!!

```{r}
# Does not really fit into the estimations from above! What have I done wrong?! I think I am misunderstanding something.
cumhazPlot(std_data$TimeUntilReinf, std_data$cens,col = 4, distr = c("wei", "loglo", "lognormal"), font.lab = 4)
```

Effect size measures: 

* lognormal: acceleration factor.
* weibull: HR.
* log-logistic: OR.

# Fit of a semi-parametric survival model

# Conclusions
